---
title: "Relationships Cheat Sheet"
author: "Nick Huntington-Klein"
date: "April 5, 2019"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning=FALSE, message=FALSE)
```

## Definitions

- **Dependent**: `X` and `Y` are *dependent* if knowing something about `X` gives you information about what `Y` is likely to be, or vice versa
- **Correlated**: `X` and `Y` are *correlated* if knowing that `X` is unusually high tells you whether `Y` is likely to be unusually high or unusually low
- **Explaining**: *Explaining* `Y` using `X` means that we are predicting what `Y` is likely to be, given a value of `X`

## Tables

`table(x)` will show us the full *distribution* of `x`. `table(x,y)` will show us the full *distribution* of `x` and `y` together (the "joint distribution"). Typically not used for continuous variables.

```{r, echo=TRUE}
library(Ecdat)
data(Benefits)

table(Benefits$joblost)

table(Benefits$joblost,Benefits$married)
```

You can label the variable names using the confusingly-named *dimnames names* option, `dnn`

```{r, echo=TRUE}
table(Benefits$joblost,Benefits$married,dnn=c('Job Loss Reason','Married'))
```

Wrap `table()` in `prop.table()` to get proportions instead of counts. The `margin` option of `prop.table()` will give the proportion within each row (`margin=1`) or within each column (`margin=2`) instead of overall.

```{r, echo=TRUE}
prop.table(table(Benefits$joblost,Benefits$married))
prop.table(table(Benefits$joblost,Benefits$married),margin=1)
prop.table(table(Benefits$joblost,Benefits$married),margin=2)
```

## Correlation

We can calculate the correlation between two (numeric) variables using `cor(x,y)`

```{r, echo=TRUE}
cor(Benefits$age,Benefits$tenure)
```

## Scatterplots

You can plot one variable against another with `plot(xvar,yvar)`. Add `xlab`, `ylab`, and `main` options to title the axes and entire plot, respectively. Use `col` to assign a color.

Use `points()` to add more points to a graph after you've made it, likely with a different `col`or.

```{r, echo=TRUE, fig.width=5, fig.height = 3.5}
library(tidyverse)
BenefitsM <- Benefits %>% filter(sex=='male')
BenefitsF <- Benefits %>% filter(sex=='female')
plot(BenefitsM$age,BenefitsM$tenure,xlab='Age',ylab='Tenure at Job',col='blue')
points(BenefitsF$age,BenefitsF$tenure,xlab='Age',ylab='Tenure at Job',col='red')
```

## Overlaid Densities

You can show how the *distribution* of `Y` changes for different values of `X` by plotting the density separately for different values of `X`. Use `lines` to add the second density plot after you've done the first one.

```{r, echo=TRUE, fig.width=5, fig.height = 3.5}
plot(density(BenefitsM$tenure),
     xlab='Tenure at Job',col='blue',main="Job Tenure by Gender")
lines(density(BenefitsF$tenure),xlab='Tenure at Job',col='red')
```

## Means Within Groups and Explaining

Part of looking at both *correlation* and *explanation* will require getting the mean of `Y` within values of `X`, which we can do with `group_by()` in `dplyr/tidyverse`.

Using `summarize()` after `group_by()` will give us a table of means within each group. Using `mutate()` will add a new variable assigning that mean. Use `mutate()` with `mean(y)` to get the part of `y` explained by `x`, or with `y - mean(y)` to get the part not explained by `x` (the residual). Don't forget to `ungroup()`!

```{r, echo=TRUE}
Benefits %>% group_by(joblost) %>% 
  summarize(tenure = mean(tenure), age = mean(age))

Benefits <- Benefits %>% group_by(joblost) %>% 
  mutate(tenure.exp = mean(tenure),
         tenure.resid = tenure - mean(tenure)) %>% ungroup()
head(Benefits %>% select(joblost,tenure,tenure.exp,tenure.resid))
```

## Explaining With a Continuous Variable

If we want to explain `Y` using `X` but `X` is continuous, we need to break it up into bins first. We will do this with `cut()`, which has the `breaks` option for how many bins to split it up into.

In this class, we will be choosing the number of breaks arbitrarily. I'll tell you what values to use.

```{r, echo=TRUE}
Benefits <- Benefits %>% mutate(agebins = cut(age,breaks=5)) %>%
  group_by(agebins) %>% 
  mutate(tenure.ageexp = mean(tenure),
         tenure.ageresid = tenure - mean(tenure)) %>% ungroup()
head(Benefits %>% select(agebins,tenure,tenure.ageexp,tenure.ageresid))
```

```{r, echo=TRUE, fig.width=5, fig.height = 3.5}
plot(Benefits$age,Benefits$tenure,xlab="Age",ylab="Tenure",col='black')
points(Benefits$age,Benefits$tenure.ageexp,col='red',cex=1.5,bg='red',pch=21)
```

## Proportion of Variance Explained

When `Y` is numeric, we can calculate its variance, and see how much of that variance is explained by `X`, and also how much is not. We do this by calculating the variance of the residuals, as this is the amount of variance in `Y` left over after taking out what `X` explains.

```{r, echo=TRUE}
#Proportion of tenure NOT explained by age
var(Benefits$tenure.ageresid)/var(Benefits$tenure)

#Proportion of tenure explained by age
1 - var(Benefits$tenure.ageresid)/var(Benefits$tenure)
var(Benefits$tenure.ageexp)/var(Benefits$tenure)
```