---
title: "Lecture 3: Introduction to R and RStudio"
author: "Nick Huntington-Klein"
date: "January 17, 2019"
output:   
  revealjs::revealjs_presentation:
    theme: solarized
    transition: slide
    self_contained: true
    smart: true
    fig_caption: true
    reveal_options:
      slideNumber: true
    
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
theme_set(theme_gray(base_size = 15))
```

## Welcome to R

The first half of this class will be dedicated to getting familiar with R.

R language for working with data. RStudio is an environment for working with that language.

By the time we're done, you should be comfortable manipulating and examining data.

## Why R?

This is going to accomplish a few things for us.

- As we mentioned last week: Excel/Sheets is a great tool for accountants, not for working with data.
- Learning to program is a highly valuable skill
- R is basically a big cool calculator, and we want to do big cool calculations

## A few notes

- These slides are written in R; you can look up the code I'm using on Titanium if you want to see how I did something
- The assigned videos are there to help
- Other resources on the [Econometrics Help](nickchk.com/econometrics.html) list, available on Titanium
- Don't be afraid of programming! It's a language like any other, just a language that requires you to be very precise. Tell the computer what you want!
- If you're already a programmer, you may get a little bored. If you'd like to learn something more advanced, let me know.
- Ask lots of questions!!! (!)

## Today

- Get used to working in RStudio
- Figure out how R conceptualizes working with stuff
- Figure out the Help system <- Important!

## The RStudio Panes

- Console
- Environment Pane
- Browser Pane
- Source Editor
- Li'l tip: Ctrl/Cmd + Shift + number will maximize one of these panes

## Console

- Typically bottom-left
- This is where you can type in code and have it run immediately
- Or, when you run code from the Source Editor, it will show up here
- It will also show any output or errors

## Console Example

- Let's copy/paste some code in there to run

```{r, echo = TRUE, eval = FALSE}
#Generate 500 heads and tails
data <- sample(c("Heads","Tails"),500,replace=TRUE)
#Calculate the proportion of heads
mean(data=="Heads")
#This line should give an error - it didn't work!
data <- sample(c("Heads","Tails"),500,replace=BLUE)
#This line should give a warning
#It did SOMETHING but maybe not what you want
mean(data)
#This line won't give an error or a warning
#But it's not what we want!
mean(data=="heads")
```

## What We Get Back

- We can see the code that we've run
- We can see the output of that code, if any
- We can see any errors or warnings (in <span style="color: red">red</span>). Remember - errors mean it didn't work. Warnings mean it *maybe* didn't work.
- Just because there's no error or warning doesn't mean it DID work! Always think carefully

## Environment Pane

- The output of what we've done can also be seen in the *Environment pane*
- This is on the top-right
- Two important tabs: Environment and History
- Mostly, Environment

## History Tab

- History shows us the commands that we've run
- To save yourself some typing, you can re-run commands by double-clicking them or hitting Enter
- Send to console with double-click/enter, send to source pane with Shift+double-click/Enter
- Or use "To Console" or "To Source" buttons

## Environment Tab

- Environment tab shows us all the objects we have in memory
- For example, we created the `data` object, so we can see that in Environment
- It shows us lots of handy information about that object too
- (we'll get to that later)
- You can erase everything with that little broom button (technically this does `rm(list=ls())`)

## Browser Pane

- Bottom-right
- Lots of handy stuff here!
- Mostly, the *outcome* of what you do will be seen here

## Files Tab

- Basic file browser
- Handy for opening up files
- Can also help you set the working directory:
    - Go to folder
    - In menu bar, Session
    - Set Working Directory
    - To Files Pane Location
  
## Plots and Viewer tabs

- When you create something that must be viewed, like a plot, it will show up here.

```{r, echo=TRUE, eval=FALSE}
data(LifeCycleSavings)
plot(density(LifeCycleSavings$pop75),main='Percent of Population over 75')
```

```{r, echo=FALSE, eval=FALSE}
## THE GGPLOT2 WAY
data(LifeCycleSavings)
ggplot(LifeCycleSavings,aes(x=pop75))+stat_density(geom='line')+
  ggtitle('Percent of Population over 75')
```

- Note the "Export" button here - this is an easy way to save plots you've created

## Packages Tab

- This is one way to install new packages and load them in
- We'll be talking more about packages later
- I generally avoid this tab; better to do this via code
- Why? Replicability! A **VERY** important reason to use code and not the GUI or, say, Excel
- You always want to make sure your future self (or someone else) knows how to use your code
- One thing I do use this for is checking for package updates, note handy "Update" button

## Help Tab

- This is where help files appear when you ask for them
- You can use the search bar here, or `help()` in the console
- We'll be going over this today a bit later

## Source Pane

- You should be working with code FROM THIS PANE, not the console!
- Why? Replicability!
- Also, COMMENTS! USE THEM! PLEASE! `#` lets you write a comment.
- Switch between tabs like a browser
- Set working directory:
    - Select a source file tab
    - In menu bar: Session
    - Set Working Directory
    - To Source File Location

## Running Code from the Source Pane

- Select a chunk of code and hit the "Run" button
- Click on a line of code and do Ctrl/Cmd-Enter to run just that line and advance to the next <- Super handy!
- Going one line at a time lets you check for errors more easily
- Let's try some!

```{r, eval=FALSE, echo=TRUE}
data(mtcars)
mean(mtcars$mpg)
mean(mtcars$wt)
372+565
log(exp(1))
2^9
(1+1)^9
```

## Autocomplete

- RStudio comes with autocomplete! 
- Typing in the Source Pane or the Console, it will try to fill in things for you
    - Command names (shows the syntax of the function too!)
    - Object names from your environment
    - Variable names in your data
- Let's try redoing the code we just did, typing it out

## Help

- Autocomplete is one way that RStudio tries to help you out
- The way that R helps you out is with the documentation
- When you start doing anything serious with a computer, like programming, the most important skills are:
     - Knowing to read documentation
     - Knowing to search the internet for help
     
## help()

- You can get the documentation on most R objects using the `help()` function
- `help(mean)`, for example, will show you:
    - What the function is
    - The "syntax" for the function
    - The available options for the function
    - Other, related functions, like `weighted.mean`
    - Ideally, some examples of proper use
- Not just for functions/commands - some data sets will work too! Try `help(mtcars)`

## Searching the Internet

- Even professional programmers will tell you - they spend a huge amount of time just looking up how to do stuff online
- Why not?
- Granted, this won't come in handy for a midterm, but for actually doing work it's essential

## Searching the Internet

- Just Google (or whatever) what you need! There will usually be a resource.
- R-bloggers, Quick-R, StackOverflow
- Or just Google. Try including "R" and the name of what you want in the search
- If "R" isn't turning up anything, try Rstats
- Ask on Twitter with `#rstats`
- Ask on StackExchange or StackOverflow <- tread lightly!

## Example

- We've picked up some data sets to use like LifeCycleSavings and mtcars with the data() function
- What data sets can we get this way?
- Let's try `help(data)`
- Let's try searching the internet

## That's it!

See you next time!