---
title: "Lecture 19 Fixed Effects"
author: "Nick Huntington-Klein"
date: "March 8, 2019"
output:   
  revealjs::revealjs_presentation:
    theme: solarized
    transition: slide
    self_contained: true
    smart: true
    fig_caption: true
    reveal_options:
      slideNumber: true
    
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning=FALSE, message=FALSE)
library(tidyverse)
library(stringr)
library(dagitty)
library(ggdag)
library(gganimate)
library(ggthemes)
library(Cairo)
library(gapminder)
library(stargazer)
theme_set(theme_gray(base_size = 15))
```

## Recap

- Last time we talked about how controlling is a common way of blocking back doors to identify an effect
- We can control for a variable `W` by using our method of using `W` to explain our other variables, then take the residuals
- Another form of controlling is using a sample that has only observations with similar values of `W`
- Some variables you want to be careful NOT to control for - you don't want to close front doors, or open back doors by controlling for colliders

## Today

- Today we'll be starting on our path for the rest of the class, where we'll be talking about standard *methods* for performing causal inference
- Different ways of getting identification once we have a diagram!
- Our goal here will be to understand these methods *conceptually*
- We won't necessarily be doing best-statistical-practices for these. You'll learn those in later classes, and best-practices change over time anyway
- Our goal is to *understand* these methods and be able to apply a straightforward version of them, not to *publish a research paper* 

## Today

- In particular we'll be talking about a method that is commonly used to identify causal effects, called fixed effects
- We'll be discussing the *kind* of causal diagram that fixed effects can identify
- All of the methods we'll be discussing are like this - they'll only apply to particular diagrams
- And so knowing our diagrams will be key to knowing when to use a given method

## The Problem

- One problem we ran into last time is that we can't really control for things if we can't measure them
- And there are lots of things we can't measure or don't have data for!
- So what can we do?

## The Solution

- If we observe each person/firm/country *multiple times*, then we can forget about controlling for the actual back-door variable we're interested in
- And just control for *person/firm/country identity* instead!
- This will control for EVERYTHING unique to that individual, whether we can measure it or not!

## In Practice

- Let's do this on the data from the "gapminder" package
- This data tracks life expectancy and GDP per capita in many countries over time

```{r, echo=TRUE, eval=FALSE}
library(gapminder)
data(gapminder)
cor(gapminder$lifeExp,log(gapminder$gdpPercap))
```
```{r, echo=FALSE, eval=TRUE}
data(gapminder)
cor(gapminder$lifeExp,log(gapminder$gdpPercap))
```

```{r, echo=TRUE}
gapminder <- gapminder %>% group_by(country) %>%
  mutate(lifeExp.r = lifeExp - mean(lifeExp),
         logGDP.r = log(gdpPercap) - mean(log(gdpPercap))) %>% ungroup()
cor(gapminder$lifeExp.r,gapminder$logGDP.r)
```

## So What?

- This isn't any different, mechanically, from any other time we've controlled for something
- So what's different here?
- Let's think about what we're doing conceptually

## What's the Diagram?

- Why are we controlling for things in this gapminder analysis?
- Because there are LOTS of things that might be back doors between GDP per capita and life expectancy
- War, disease, political institutions, trade relationships, health of the population, economic institutions...

## What's the Diagram?

```{r, dev='CairoPNG', echo=FALSE, fig.width=5, fig.height=6}
dag <- dagify(LifeEx~GDPpc+A+B+C+D+E+F+G+H,
              GDPpc~A+B+C+D+E+F+G+H,
              coords=list(
                x=c(LifeEx=4,GDPpc=2,A=1,B=2,C=3,D=4,E=1,F=2,G=3,H=4),
                y=c(LifeEx=2,GDPpc=2,A=3,B=3,C=3,D=3,E=1,F=1,G=1,H=1)
              )) %>% tidy_dagitty()
ggdag(dag,node_size=20)
```

## What's the Diagram?

- There's no way we can identify this
- The list of back doors is very long
- And likely includes some things we can't measure!

## What's the Diagram?

- HOWEVER! If we think that these things are likely to be constant within country...
- Then we don't really have a big long list of back doors, we just have one: "country"

```{r, dev='CairoPNG', echo=FALSE, fig.width=5, fig.height=3.5}
dag <- dagify(LifeEx~GDPpc+Coun,
              GDPpc~Coun,
              coords=list(
                x=c(LifeEx=4,GDPpc=2,Coun=3),
                y=c(LifeEx=2,GDPpc=2,Coun=3)
              )) %>% tidy_dagitty()
ggdag(dag,node_size=20)
```

## What We Get

- So what we get out of this is that we can identify our effect even if some of our back doors include variables that we can't actually measure
- When we do this, we're basically comparing countries *to themselves* at different time periods!
- Pretty good way to do an apples-to-apples comparison!

## Graphically

```{r, dev='CairoPNG', echo=FALSE, fig.width=7, fig.height=5}
lgdpmean <- mean(log(gapminder$gdpPercap))
lemean <- mean(gapminder$lifeExp)
ggplot(gapminder)+geom_point(aes(x=log(gdpPercap)-lgdpmean,y=lifeExp-lemean,color="Raw"),alpha=.6)+
  geom_point(aes(x=logGDP.r,y=lifeExp.r,color="After Fixed Effects"))+
  geom_point(data=filter(gapminder,country=="Pakistan"),
             aes(x=log(gdpPercap)-lgdpmean,y=lifeExp-lemean,color="Raw Pakistan"),size=3)+
  labs(x="log(GDP Per Capita)",y="Life Expectancy")+
  theme(axis.text.x=element_blank(),
        axis.text.y=element_blank())+
  scale_color_manual(name="Values",values=c("After Fixed Effects" = "blue","Raw" = "black","Raw Pakistan" = "red"))
```

## Graphically

- The post-fixed-effects dots are basically a bunch of "Raw Country X" pasted together.
- Imagine taking "Raw Pakistan" and moving it to the center, then taking "Raw Britain" and moving it to the center, etc.
- Ignoring the baseline differences between Pakistan, Britain, China, etc., in their GDP per capita and life expectancy, and just looking within each country.
- We are ignoring all differences *between* countries (since that way back doors lie!) and looking only at differences *within* countries. 
- Fixed Effects is sometimes also referred to as the "within" estimator

## In Action

```{r, dev='CairoPNG', echo=FALSE, fig.width=7, fig.height=5}
df <- data.frame(Person = rep(1:4,50)) %>%
  mutate(X = .5+.5*(Person-2.5) + rnorm(200)) %>%
  mutate(Y = -.5*X + (Person-2.5) + 1 + rnorm(200),time="1") %>%
  group_by(Person) %>%
  mutate(mean_X=mean(X),mean_Y=mean(Y)) %>%
  ungroup()

#Calculate correlations
before_cor <- paste("1. Start with raw data. Correlation between X and Y: ",round(cor(df$X,df$Y),3),sep='')
after_cor <- paste("6. Within-Individual Correlation Between X and Y: ",round(cor(df$X-df$mean_X,df$Y-df$mean_Y),3),sep='')

#Add step 2 in which X is demeaned, and 3 in which both X and Y are, and 4 which just changes label
dffull <- rbind(
  #Step 1: Raw data only
  df %>% mutate(mean_X=NA,mean_Y=NA,time=before_cor),
  #Step 2: Add x-lines
  df %>% mutate(mean_Y=NA,time='2. Figure out between-Individual differences in X'),
  #Step 3: X de-meaned 
  df %>% mutate(X = X - mean_X,mean_X=0,mean_Y=NA,time="3. Remove all between-Individual differences in X"),
  #Step 4: Remove X lines, add Y
  df %>% mutate(X = X - mean_X,mean_X=NA,time="4. Figure out between-Individual differences in Y"),
  #Step 5: Y de-meaned
  df %>% mutate(X = X - mean_X,Y = Y - mean_Y,mean_X=NA,mean_Y=0,time="5. Remove all between-Individual differences in Y"),
  #Step 6: Raw demeaned data only
  df %>% mutate(X = X - mean_X,Y = Y - mean_Y,mean_X=NA,mean_Y=NA,time=after_cor))

p <- ggplot(dffull,aes(y=Y,x=X,color=as.factor(Person)))+geom_point()+
  geom_vline(aes(xintercept=mean_X,color=as.factor(Person)))+
  geom_hline(aes(yintercept=mean_Y,color=as.factor(Person)))+
  guides(color=guide_legend(title="Individual"))+
  scale_color_colorblind()+
  labs(title = 'The Relationship between Y and X, with Individual Fixed Effects \n{next_state}')+
  transition_states(time,transition_length=c(12,32,12,32,12,12),state_length=c(160,100,75,100,75,160),wrap=FALSE)+
  ease_aes('sine-in-out')+
  exit_fade()+enter_fade()

animate(p,nframes=200)
```

## Notably

- This does assume, of course, that all those back door variables CAN be described by country
- In other words, that these back doors operate by things that are *fixed* within country
- If something is a back door and changes over time in that country, fixed effects won't help!

## Varying Over Time

- For example, earlier we mentioned war... that's not fixed within country! A given country is at war sometimes and not other times.

```{r, dev='CairoPNG', echo=FALSE, fig.width=5, fig.height=4}
dag <- dagify(LifeEx~GDPpc+Coun+War,
              GDPpc~Coun+War,
              coords=list(
                x=c(LifeEx=4,GDPpc=2,Coun=3.5,War=2.5),
                y=c(LifeEx=2,GDPpc=2,Coun=3,War=3)
              )) %>% tidy_dagitty()
ggdag(dag,node_size=20)
```

## Varying Over Time

- Of course, in this case, we could control for War as well and be good!
- Time-varying things doesn't mean that fixed effects doesn't work, it just means you need to control for that stuff too
- It always comes down to thinking carefully about your diagram
- Fixed effects mainly works as a convenient way of combining together lots of different constant-within-country back doors into something that lets us identify the model even if we can't measure them all

## Example: Sentencing

- What effect do sentencing reforms have on crime? 
- One purpose of punishment for crime is to deter crime
- If sentences are more clear and less risky, that may reduce a deterrent to crime and so increase crime
- Marvell & Moody study this using data on reforms in US states from 1969-1989

## Example: Sentencing

```{r, echo=FALSE}
mm <- as.data.frame(readLines('marvel_moody_sentencing.txt'))
mm1 <- as.data.frame(mm[rep(c(TRUE,FALSE),2100/2),]) %>%
  rename(mm1 = `mm[rep(c(TRUE, FALSE), 2100/2), ]`)
mm2 <- as.data.frame(mm[rep(c(FALSE,TRUE),2100/2),]) %>%
  rename(mm2 = `mm[rep(c(FALSE, TRUE), 2100/2), ]`)
mmdata <- tibble(
  state = substr(mm1$mm1,5,8),
  year = as.numeric(substr(mm1$mm1,11,12)),
  assault = as.numeric(substr(mm1$mm1,44,49)),
  robbery = as.numeric(substr(mm1$mm1,50,55)),
  pop1000 = as.numeric(substr(mm1$mm1,56,61)),
  sentreform = as.numeric(str_sub(trimws(mm2$mm2),-3))
) %>%
  mutate(sentreform = ceiling(sentreform)) %>%
  na.omit
```

- I've omitted code reading in the data
- But in our data we have multiple observations per state

```{r, echo=TRUE}
head(mmdata)
mmdata <- mmdata %>% mutate(assaultper1000 = assault/pop1000,
         robberyper1000 = robbery/pop1000)
```

## Fixed Effects

- We can see how robbery rates evolve in each state over time as states implement reform

```{r, dev='CairoPNG', echo=FALSE, fig.width=8, fig.height=4.5}
ggplot(mmdata,aes(x=year,y=robberyper1000,
                group=state,color=factor(sentreform)))+
  geom_line(size=1)+scale_color_colorblind(name="Reform")+
  labs(x="Year",y="Robberies per 1000 Population")
```

## Fixed Effects

- You can tell that states are more or less likely to implement reform in a way that's correlated with the level of robbery they already had
- So SOMETHING about the state is driving both the level of robberies AND the decision to implement reform
- Who knows what!
- Our diagram has `reform -> robberies` and `reform <- state -> robberies`, which is something we can address with fixed effects.

## Fixed Effects

```{r, echo=TRUE}
cor(mmdata$sentreform,mmdata$robberyper1000)
mmdata <- mmdata %>% group_by(state) %>%
  mutate(reform.m = sentreform-mean(sentreform),
         robbery.m = robberyper1000-mean(robberyper1000))
cor(mmdata$reform.m,mmdata$robbery.m)
```

## Example

- The numbers were different
- The `r round(cor(mmdata$sentreform,mmdata$robberyper1000),3)` included the fact that different kinds of states tend to institute reform
- The `r round(cor(mmdata$reform.m,mmdata$robbery.m),3)` doesn't!
- Looks like the deterrent effect was real! Although important to consider if there might be time-varying back doors too, we don't account for those in our analysis
- What things might change within state over time that would be related to robberies and to sentencing reform?

## Practice

- We want to know the effect of your `teacher` on the `test score`s of high school students
- Some potential back doors might go through: `parents' intelligence`, `age`, `demographics`, `school`, `last year's teacher`
- Draw a diagram including all these variables, plus maybe some unobservables where appropriate
- If you used fixed effects for students, what back doors would still be open?

## Practice Answers

- Fixed effects would close your back doors for `parents' intelligence`, `demographics`, and `school`, but leave open `age` and `last year's teacher`