---
title: "Lecture 4: Objects and Functions in R"
author: "Nick Huntington-Klein"
date: "January 17, 2019"
output:   
  revealjs::revealjs_presentation:
    theme: solarized
    transition: slide
    self_contained: true
    smart: true
    fig_caption: true
    reveal_options:
      slideNumber: true
    
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
theme_set(theme_gray(base_size = 15))
```

## Working in R

- Today we'll be starting to actually DO stuff in R
- R is all about:
    - Creating objects
    - Looking at objects
    - Manipulating objects
    - That's, uh, it. That's all. That's what R does.

## Creating a Basic Object

- Let's create an object. We're going to do this with the assignment operator `<-` (a.k.a. "gets") 

```{r, echo=TRUE}
a <- 4
```
- This creates an object called `a`. What is that object? It's a number. Specifically, it's the number 4. We know that because we took that 4 and we shoved it into `a`.
- Why store it as an object rather than just saying 4? Oh, plenty of reasons.
- We can do more complex calculations before storing it, too.
```{r, echo=TRUE}
b <- sqrt(16)+10
```


## Looking at Objects

- We can see this object that we've created in the Environment pane
- Putting that object on a line by itself in R will show us what it is.

```{r, echo=TRUE}
a
b
```

## Looking at Objects 

- We can even create an object and look at it in the console without storing it.
- Let's think about what is really happening in these lines
```{r, echo=TRUE}
3
a+b
```

## Looking at Objects Differently

- We can run objects through **functions** to look at them in different ways

```{r, echo=TRUE}
#What does a look like if we take the square root of it?
sqrt(a)
#What does it look like if we add 1 to it?
a + 1
#If we look at it, do we see a number?
is.numeric(a)
```

## Manipulating Objects

- We manipulated the object and LOOKED at it, we can also SAVE it, or UPDATE it.
```{r, echo=TRUE}
#We looked at what a looked like with 1 added, but a itself wasn't changed
a
#Let's save a+1 as something else
b <- a + 1
#And let's overwrite a with its square root
a <- sqrt(a)
a
b
```

## Some Notes

- Even though we changed `a`, `b` was already set using the old value of `a`, and so was still `4+1=5`, not `2+1=3`. 
- `a` basically got *reassigned* with `<-`. That's how we got it to be `2`
- This is a very simple example, but basically *everything* in R is just this process but with more complex objects and more complex functions!

## Types of Objects

- We already determined that `a` was a number. But what else could it be? What other kinds of variables are there?
- Some basic object types:
    - Numeric: A single number
    - Character: A string of letters, like `'hello'`
    - Logical: `TRUE` or `FALSE` (or `T` or `F`)
    - Factor: A category, like `'left handed', 'right handed', or 'ambidextrous'`
    - Vector: A collection of objects of the same type
    
## Characters

- A character object is a piece of text, held in quotes like `''` or `""`
- For example, maybe you have some data on people's addresses.

```{r, echo=TRUE}
address <- '321 Fake St.'
address
is.character(address)
```

## Logical

- Logicals are binary - TRUE or FALSE. They're extremely important because lots of data is binary.

```{r, echo=TRUE}
c <- TRUE
is.logical(c)
is.character(a)
is.logical(is.numeric(a))
```

## Logical

- Logicals are used a lot in programming too, because you can evaluate whether conditions hold.

```{r, echo=TRUE}
a > 100
a > 100 | b == 5
```

- `&` is AND, `|` is OR, and to check equality use `==`, not `=`. `>=` is greater than OR equal to, similarly for `<=`

## Logical

- They are also equivalent to `TRUE=1` and `FALSE=0` which comes in handy.
- We can use `as` functions to change one object type to another (although for `T=1, F=0` it does it automatically)

```{r, echo=TRUE}
as.numeric(FALSE)
TRUE + 3
```

## Logicals Test

- Let's stop and try to think about what these lines might do:
```{r, echo=TRUE, eval=FALSE}
is.logical(is.numeric(FALSE))
is.numeric(2) + is.character('hello')
is.numeric(2) & is.character(3)
TRUE | FALSE
TRUE & FALSE
```

## Factors

- Factors are categorical variables - mutually exclusive groups
- They look like strings, but they're more like logicals, but with more than two levels (and labeled differently than T and F)
- Factors have *levels* showing the possible categories you can be in
```{r, echo=TRUE}
e <- as.factor('left-handed')
levels(e) <- c('left-handed','right-handed','ambidextrous')
e
```

## Vectors

- Data is basically a bunch of variables all put together
- So unsurprisingly, a lot of R works with vectors, which are a bunch of objects all put together!
- Use `c()` (concatenate) to put a bunch of objects of the same type together in a vector
- Use square brackets to pick out parts of the vector

```{r, echo=TRUE}
d <- c(5,6,7,8)
c(is.numeric(d),is.vector(d))
d[2]
```

## Vectors

- Unsurprisingly, lots of statistical functions look at multiple objects (what is statistics but making sense of lots of different measurements of the same thing?)

```{r, echo=TRUE}
mean(d)
c(sum(d),sd(d),prod(d))
```

## Vectors

- We can perform the same operation on all parts of the vector at once!

```{r, echo=TRUE}
d + 1
d + d
d > 6
```

## Vectors

- Factors make a lot more sense as a vector

```{r, echo=TRUE}
continents <- as.factor(c('Asia','Asia','Asia',
                          'N America','Europe','Africa','Africa'))
table(continents)
continents[4]
```

## Value Matching

- It's easy to create logicals seeing if a value matches ANY value in a vector with %in%

```{r, echo=TRUE}
3 %in% c(3,4)
c('Nick','James') %in% c('James','Andy','Sarah')
```

## Basic Vectors

- Generating basic vectors:

```{r, echo=TRUE, eval=FALSE}
1:8
rep(4,3)
rep(c('a','b'),4)
numeric(5)
character(6)
sample(1:20,3)
sample(c("Heads","Tails"),6,replace=TRUE)
```
```{r, echo=FALSE, eval=TRUE}
1:8
rep(4,3)
rep(c('a','b'),4)
numeric(5)
character(6)
sample(1:20,3)
sample(c("Heads","Tails"),6,replace=TRUE)
```

## Vector Test
- If we do `f <- c(2,3,4,5)`, then what will the output of these be?

```{r, echo=TRUE, eval = FALSE}
f^2
f + c(1,2,3,4)
c(f,6)
is.numeric(f)
mean(f >= 4)
f*c(1,2,3)
length(f)
length(rep(1:4,3))
f/2 == 2 | f < 3
as.character(f)
f[1]+f[4]
c(f,f,f,f)
f[f[1]]
f[c(1,3)]
f %in% (1:4*2)
```

## And Now You

- Create a factor that randomly samples six `'Male'` or `'Female'` people.
- Add up all the numbers from 18 to 763, then get the mean
- What happens if you make a list with a logical, a numeric, AND a string in it?
- Figure out how to use `paste0()` to turn `c('a','b')` into `'ab'`
- Use `[]` to turn `h <- c(10,9,8,7)` into `c(7,8,10,9)` and call it `j`
- (Several ways) Create a vector with eleven 0's, then a 5.
- (Tough!) Use `floor()` or `%%` to count how many multiples of 4 there are between 433 and 899